Closing the empirical cycle

### 2022-06-09 presentation_vvsor.Rmd
add vvs_or presentation

---
title: "Developmental Data Science"
author: "Caspar J. Van Lissa<sup>1</sup>"
date: "<p><sup>1</sup>: Utrecht University, dept. Methodology & Statistics | Open Science Community Utrecht<br>Funded by NWO Veni grant  VI.Veni.191G.090"
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "docs") })
output: 
  ioslides_presentation:
    widescreen: true
    css: slide.css
bibliography      : ["manuscript/references.bib", "../rigorous_exploration/manuscript/r-references.bib", "../rigorous_exploration/manuscript/references.bib", "../rigorous_exploration/manuscript/Rigorous_exploration.bib", "../rigorous_exploration/manuscript/Veni.bib", "../rigorous_exploration/manuscript/Veni_sysrev.bib"]
---

@@

Why do we need theory?

* Useful as an instrument of cumulative knowledge acquisition
* Communicating social scientific research to its consumers

@@

@@

@@

## Machine learning vs unguided exploration

```{r}
knitr::include_graphics("Figure Machine learning 4.png")
```

@@

@@

## Knowledge gap

Substantial empirical research, but:

* Different (sub)disciplines have approached the topic in disparate ways [@riedigerEmotionRegulationAdolescence2014]
* No consistent terminology [@bariolaChildAdolescentEmotion2011]
* No conceptual frameworks [@stifterEmotionRegulation2019]
* No overarching theoretical framework [@bussTheoriesEmotionalDevelopment2019]

Knowledge must be consolidated into overarching theory

## Limitations of relevant theory

* Few explicitly address adolescence
* Few comprehensively address predictors of emotion regulation
* None directly guide contemporary research 
* Theories vary widely in scope: Some are broad and non-specific; others describe a specific phenomenon in detail, but lack a broader perspective.
    + Broad theories can frame any research, specific theories generate hypotheses.
    + It would be beneficial to bridge these levels of analysis.

@@

## Baseline

Manual selection of key constructs from relevant theory and prior reviews [ @bariolaChildAdolescentEmotion2011; @coe-odessEmergentEmotionsAdolescence2019]:

* Bioecological model [@bronfenbrennerBioecologicalModelHuman2007]
* Transactional model [@sameroffUnifiedTheoryDevelopment2010]
* Hall's notion of "storm and stress"
* Theory of normative emotional development [@sroufeEmotionalDevelopmentOrganization1995]
* Tripartite model [@morrisRoleFamilyContext2007]
* Internalization model [@holodynskiDevelopmentEmotionsEmotion2006]
* Polyvagal theory [@porgesOrientingDefensiveWorld1995].
* Model of social-affective engagement and goal flexibility [@croneUnderstandingAdolescencePeriod2012]
* Process model of emotion regulation [@grossHandbookEmotionRegulation2013]
* Social Information Processing Theory [@lemeriseIntegratedModelEmotion2000]

@@

## {.fullslide}

![](../veni_sysrev/varimps_small.png)

## Co-occurrence graph

@@

## Results

@@

* Developmental disorders
* Physical health (sic)
* External stressors
* Structural disadvantage
* Addictive behavior
* Identity and moral development
* Sexual development

## Implications

* Empirical research relies on theory; under-theorized phenomena may be overlooked
    + This study offers guidance for phenomena to consider as confounders/causes
* Unembedded terms indicate promising areas of future research
* Present study serves as starting point for theory development

## Next step

@@

```{r}
knitr::include_graphics("tree.png")
```

@@

* RADAR data (https://www.uu.nl/en/research/radar)
* 497 Dutch adolescents (283 boys; age at T1: M =13.03, SD = 0.46)
* Collected between 2006-2011
* Most families were  medium- to high-SES (10% low-SES)

## Candidate predictors

Predictors: 87

Categories:

* Demographics (e.g., SES)
* Biological (age, sex, pubertal development)
* Individual differences (big5, BIS/BAS, empathy)
* Risk behavior (in/externalizing symptoms, substance use, delinquency, school performance, victimization)
* Relationship quality (parents/peers)
* Parenting (perceived and parent-reported)
* Conflict resolution styles

## Dependent variable

Difficulties in emotion regulation [@gratzMultidimensionalAssessmentEmotion2004]

* 24 items (omitted 'emotional awareness' subscale)
* E.g.: “When I’m upset, I have difficulty thinking about anything else”
* 1 (“Almost never”) to 5 (“Almost always”)
* Excellent reliability, $(.94 > \alpha > .95)$.
* Metric invariance across waves
* Quadratic growth model had good fit
  + CFI = .996, TLI = .993, RMSEA = .047, SRMR = 0.020

@@

```{r, out.width="40%"}
knitr::include_graphics("plots/pdp1-12.png")
```

Important:

* Best predictors: Personality and related constructs
* Best parental predictor is autonomy support (bal. rel.)
  + Negative/intrusive parenting more predictive than other positive practices
* Conflict frequency and behavior
* Empathy, particularly personal distress

>>>>

---
title: "Closing the empirical cycle"
author: "Using machine learning for rigorous exploration"
date: "Caspar J. Van Lissa<p>Utrecht University | Open Science Community Utrecht<br>Tilburg University<br>Funded by NWO Veni grant  VI.Veni.191G.090</p>"
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "docs") })
output: 
  ioslides_presentation:
    widescreen: true
    css: slide.css
bibliography      : ["manuscript/references.bib", "../rigorous_exploration/manuscript/r-references.bib", "../rigorous_exploration/manuscript/references.bib", "../rigorous_exploration/manuscript/Rigorous_exploration.bib", "../rigorous_exploration/manuscript/Veni.bib", "../rigorous_exploration/manuscript/Veni_sysrev.bib"]
---

@@

Why should statisticians care about theory?

* It dictates what models are sensible
* Useful as an instrument of cumulative knowledge acquisition
* Communicating social scientific research to its consumers

@@

$$
FDR = \frac{FP}{TP + FP}
$$

@@

* E.g., personalized (mental) health care, automated assessment aids

@@

## Unguided vs rigorous exploration

```{r, out.width="100%"}
knitr::include_graphics("unguidedvrigorous.png")
```

@@

Substantial empirical research, but no overarching theoretical framework [@bussTheoriesEmotionalDevelopment2019]

@@

@@

@@

## TMSR results

@@

## Interpretation

@@

* Developmental disorders
* Physical health
* External stressors
* Structural disadvantage
* Addictive behavior
* Identity and moral development
* Sexual development

## Next step: Identifying individual risk factors

@@

```{r, out.width = "65%"}
knitr::include_graphics("tree.png")
```

@@

* RADAR data (https://www.uu.nl/en/research/radar)
* 497 Dutch adolescents (283 boys; age at T1: M =13.03, SD = 0.46)
* Collected between 2006-2011
* Most families were  medium- to high-SES (10% low-SES)
* 87 candidate predictors
* DV: quadratic growth model of Difficulties in emotion regulation [@gratzMultidimensionalAssessmentEmotion2004]

@@

```{r, out.width="37%"}
knitr::include_graphics("plots/pdp1-12.png")
```

## Interpretation

Important predictors:

* Personality and related constructs
* Best parental predictor is autonomy support (bal. rel.)
  + Negative parenting more predictive than positive parenting
* Conflict frequency and behavior
* Empathy, particularly personal distress
